---
title: "Cohort analysis"
---
<style>
    body .main-container {
        max-width: 2000px;
    }
</style>

# Introduction

This R notebook is a rapid analysi of the cohort data produced as part of the data synthesis of citizen science projects

```{r setup, echo=FALSE, results=FALSE, message=FALSE}
library(plotly)
library(readxl)
library(tidyr)
library(reshape2)

knitr::opts_chunk$set(out.width = "80%", fig.align="center")
```

# Data

Lets load in the data, it is currently seperate sheets in an excel file. Lets pull out the seperate sheets and assign to `data.frame`s

```{r load_data, message=FALSE, echo=FALSE}
summary <- read_excel(path = 'data/BES citsci data summary v3.xlsx', sheet = 1)
metadata <- read_excel(path = 'data/BES citsci data summary v3.xlsx', sheet = 2)
retention_tasks <- read_excel(path = 'data/BES citsci data summary v3.xlsx', sheet = 3)
retention_time <- read_excel(path = 'data/BES citsci data summary v3.xlsx', sheet = 4)
ppt_inequality <- read_excel(path = 'data/BES citsci data summary v3.xlsx', sheet = 5)
```

I think I need to spread the cohort data so that time is represented in columns

```{r, echo=FALSE}
ret <- pivot_wider(data = retention_time, names_from = 'Session', values_from = 'NumberOfPeople')
head(ret)

# easy to view but not he format for ggplot

retention_time$CohortSession <- (retention_time$Session + 1) - retention_time$Cohort 
retention_time$ProjectCohort <- paste(retention_time$ProjectCode, 
                                      retention_time$Cohort,
                                      sep = '_')
retention_time$ProjectName <- NA
retention_time$ProjectName <- summary$ProjectName[match(retention_time$ProjectCode, summary$ProjectCode)]
retention_time$ProjectName[is.na(retention_time$ProjectName)] <-
  as.character(retention_time$ProjectCode[is.na(retention_time$ProjectName)])
```

That works. So now can we do an initial visualisation of the data

```{r, echo=FALSE}
p <- ggplot(retention_time, aes(x = CohortSession, y = NumberOfPeople, group = ProjectCohort)) +
  geom_line(aes(colour = ProjectName)) +
  theme(legend.position = "none") +
  ggtitle('Raw data')
ggplotly(p, tooltip = c('group'))

```

Okay now we need to average across cohorts in the same project and rescale so they all start with the same value

```{r, echo=FALSE}
# Rescale
rescale <- function(x){
  
  (x/max(x)) * 100
  
}

retention_time$NumberOfPeopleRescaled <- NA

for(i in unique(retention_time$ProjectCohort)){
  
  retention_time$NumberOfPeopleRescaled[retention_time$ProjectCohort == i] <-
    rescale(retention_time$NumberOfPeople[retention_time$ProjectCohort == i])
  
}


p <- ggplot(retention_time, aes(x = CohortSession, y = NumberOfPeopleRescaled, group = ProjectCohort)) +
  geom_line(aes(colour = ProjectName)) +
  theme(legend.position = "none")  +
  ggtitle('Rescaled cohorts')
ggplotly(p, tooltip = c('group'))

```
And now average across projects

```{r, echo=FALSE}
# group and average
av_ret <- tapply(retention_time$NumberOfPeopleRescaled,
                 INDEX = list(retention_time$ProjectCode,
                              retention_time$CohortSession),
                 FUN = mean)

# Put into long format
av_ret <- melt(av_ret, varnames = c('ProjectCode', 'CohortSession'), value.name = 'NumberOfPeopleRescaled')

av_ret$ProjectName <- summary$ProjectName[match(av_ret$ProjectCode, summary$ProjectCode)]
av_ret$ProjectName[is.na(av_ret$ProjectName)] <-
  as.character(av_ret$ProjectCode[is.na(av_ret$ProjectName)])

p <- ggplot(av_ret, aes(x = CohortSession, y = NumberOfPeopleRescaled,
                        group = ProjectName)) +
  geom_line(aes(colour = ProjectName)) +
  geom_point(aes(colour = ProjectName)) +
  theme(legend.position = "none")  +
  ggtitle('Project averages')
ggplotly(p, tooltip = c('group'))

```

After the first year these actually look very similar

```{r, echo=FALSE}
retention_time$NumberOfPeopleRescaled <- NA

retention_time <- retention_time[retention_time$CohortSession != 1,]

for(i in unique(retention_time$ProjectCohort)){
  
  retention_time$NumberOfPeopleRescaled[retention_time$ProjectCohort == i] <-
    rescale(retention_time$NumberOfPeople[retention_time$ProjectCohort == i])
  
}

# group and average
av_ret2 <- tapply(retention_time$NumberOfPeopleRescaled,
                 INDEX = list(retention_time$ProjectCode,
                              retention_time$CohortSession),
                 FUN = mean)

# Put into long format
av_ret2 <- melt(av_ret2, varnames = c('ProjectCode', 'CohortSession'), value.name = 'NumberOfPeopleRescaled')

av_ret2$ProjectName <- 
  summary$ProjectName[match(av_ret2$ProjectCode,
                            summary$ProjectCode)]
av_ret2$ProjectName[is.na(av_ret2$ProjectName)] <-
  as.character(av_ret2$ProjectCode[is.na(av_ret2$ProjectName)])

p <- ggplot(av_ret2, aes(x = CohortSession, y = NumberOfPeopleRescaled,
                        group = ProjectName)) +
  geom_line(aes(colour = ProjectName)) +
  geom_point(aes(colour = ProjectName)) +
  theme(legend.position = "none")  +
  ggtitle('Project averages from year 2')
ggplotly(p, tooltip = c('group'))
```
And on the log scale...


```{r, echo=FALSE}
p <- ggplot(av_ret2, aes(x = CohortSession, y = NumberOfPeopleRescaled,
                        group = ProjectName)) +
  geom_line(aes(colour = ProjectName)) +
  geom_point(aes(colour = ProjectName)) +
  scale_y_log10() +
  theme(legend.position = "none")  +
  ggtitle('Project averages from year 2 - log10 scale')
ggplotly(p, tooltip = c('group'))
```

```{r, echo=FALSE}
met <- as.data.frame(t(metadata))
colnames(met) <- met[1,]
met$ProjectCode <- rownames(met)
met <- met[2:nrow(met), -10]
# names(met)
av_ret3 <- merge(av_ret2, met, by = 'ProjectCode', all.x = TRUE, all.y = FALSE)
```

What is your target audience?
* Wide audience (i.e. 'anyone', general public, interested 'novice')
* Specific audience (e.g. farmers, birdwatchers)

```{r, echo=FALSE}
# TargetAudience
p <- ggplot(av_ret3, aes(x = CohortSession, y = NumberOfPeopleRescaled,
                        group = ProjectCode)) +
  geom_line(aes(colour = TargetAudience)) +
  geom_point(aes(colour = TargetAudience)) +
  ggtitle('TargetAudience') + 
  theme(legend.title = element_blank())

ggplotly(p, tooltip = c('group'))
```

What experience or skill is required to take part?

* None: anyone can do it by following simple instructions
* Expertise is required but training is given (supporting materials etc.)
* Pre-existing expertise is required (e.g. skill in identifying birds)

```{r, echo=FALSE}
# ExperienceRequired
p <- ggplot(av_ret3, aes(x = CohortSession, y = NumberOfPeopleRescaled,
                        group = ProjectCode)) +
  geom_line(aes(colour = ExperienceRequired)) +
  geom_point(aes(colour = ExperienceRequired)) +
  ggtitle('ExperienceRequired') + 
  theme(legend.title = element_blank())

ggplotly(p, tooltip = c('group'))
```

What effort is required to take part?

* Quick and easy = little or no planning required to take part
* Specific planning = requires time, special equipment and/or travel to a particular site

```{r, echo=FALSE}
# EffortRequired
p <- ggplot(av_ret3, aes(x = CohortSession, y = NumberOfPeopleRescaled,
                        group = ProjectCode)) +
  geom_line(aes(colour = EffortRequired)) +
  geom_point(aes(colour = EffortRequired)) +
  ggtitle('EffortRequired') + 
  theme(legend.title = element_blank())

ggplotly(p, tooltip = c('group'))
```

Does participation require visits to pre-defined locations?

* Yes
* No

```{r, echo=FALSE}
# LocationsPredefined
p <- ggplot(av_ret3, aes(x = CohortSession, y = NumberOfPeopleRescaled,
                        group = ProjectCode)) +
  geom_line(aes(colour = LocationsPredefined)) +
  geom_point(aes(colour = LocationsPredefined)) +
  ggtitle('LocationsPredefined') + 
  theme(legend.title = element_blank())

ggplotly(p, tooltip = c('group'))
```

Is feedback given at or near the time of submission?

* No
* Yes, general
* Yes, detailed and/or personal

```{r, echo=FALSE}
# SubmissionFeedback
p <- ggplot(av_ret3, aes(x = CohortSession, y = NumberOfPeopleRescaled,
                        group = ProjectCode)) +
  geom_line(aes(colour = SubmissionFeedback)) +
  geom_point(aes(colour = SubmissionFeedback)) +
  ggtitle('SubmissionFeedback') + 
  theme(legend.title = element_blank())

ggplotly(p, tooltip = c('group'))
```

Is static feedback given (e.g. reports at the end of or at key times within the project)? 

* Yes
* No

```{r, echo=FALSE}
# SummaryFeedback
p <- ggplot(av_ret3, aes(x = CohortSession, y = NumberOfPeopleRescaled,
                        group = ProjectCode)) +
  geom_line(aes(colour = SummaryFeedback)) +
  geom_point(aes(colour = SummaryFeedback)) +
  ggtitle('SummaryFeedback') + 
  theme(legend.title = element_blank())

ggplotly(p, tooltip = c('group'))
```

In addition to field-based recording, are there other ways for your target audience to contribute within the project?

* Yes
* No

```{r, echo=FALSE}
# OtherWaysToContribute
p <- ggplot(av_ret3, aes(x = CohortSession, y = NumberOfPeopleRescaled,
                        group = ProjectCode)) +
  geom_line(aes(colour = OtherWaysToContribute)) +
  geom_point(aes(colour = OtherWaysToContribute)) +
  ggtitle('OtherWaysToContribute') + 
  theme(legend.title = element_blank())

ggplotly(p, tooltip = c('group'))
```